Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-03 paper/paper.tex
garbage collection --> GC

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis  has  two parts.   We  first  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands  on its  arguments.   This  summary is  used  to step  through
function calls  during analysis.   The concrete  demand on  a function
body  is  obtained through  a  conservative  approximation similar  to
0-CFA~\cite{Shivers:1988} that  combines the demands on  all the calls
to  the function.   The result  of the  analysis is  an annotation  of
certain program points with  deterministic finite-state automata (DFA)
capturing the  liveness of  variables at  these points.   Depending on
where garbage collection is triggered, the collector consults a set of
automata to restrict reachability during  marking.  This results in an
increase in  the garbage reclaimed  and consequently in  fewer garbage
collections.

\todelete{While  this  idea has  been  shown  to  be effective  for  a
  first-order     {\em     eager}    language~\cite{asati14lgc},     a
  straightforward extension of the technique  is not possible for lazy
  languages,  where   heap-allocated  objects  may   include  closures
  (runtime   representations    of   unevaluated    expressions).    }
\added{Designing a liveness based garbage collector for lazy languages
  poses significant challenges.}  Firstly, since  data is made live by
evaluation of closures, and in lazy languages the place in the program
where  this evaluation  takes place  cannot be  statically determined,
laziness   complicates  liveness   analysis  itself.    Moreover,  for
liveness-based garbage collection  to be effective, we  need to extend
it to closures apart from evaluated  data.  Since a closure can escape
the  scope in  which it  was  created, it  has to  carry the  liveness
information  of  its  free  variables.  As  execution  progresses  and
possible  future   uses  are   eliminated,  we  update   the  liveness
information  in a  closure with  a more  precise version.   Because of
these reasons,  the garbage collector also  becomes significantly more
complicated than a liveness-based collector for a eager language.

@@

%------------------------------------------------------------%
\subsection{Motivating example}
\label{sec:motiv}
\input{fig_motiv}   Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e. just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus if a garbage collection takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are
\added{
\begin{itemize}
\item We propose a liveness based garbage collection scheme for a lazy
  first order functional  language and prove its  correctness.  To the
  best  of our  knowledge, this  is the  first work  on liveness-based
  garbage    collection   for    lazy    languages.    All    previous
  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
    asati14lgc,    karkare06effectiveness}   have    involved   either
  imperative  or  eager  functional  languages.   The  novely  of  our
  approach is in  handling of  liveness of  functional values  (i.e.,
  closures), an issue that arises due to laziness.

@@

\item The  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether to  copy a  cell during garbage  collection translates  to a
  membership problem for such grammars.   In this paper, we have shown
  the undecidability  of this  problem. We  circumvent the  problem by
  over-approximating the grammars with DFAs.

\item We have implemented a garbage  collector that uses the result of
  liveness  analysis to  retain  live cells.   Our experiments  reveal
  interesting  space-time   trade-offs  in  the  engineering   of  the
  collector---for  example, updating  liveness information  carried in
  closures during  execution results in more  garbage being collected.
  Empirical results  show the effectiveness of  liveness-based garbage
  collection.
\end{itemize}
}
\todelete{
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
  functional language and  prove its correctness. The 
  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
  correct with respect to the specification.   
\item The analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether   to  copy  a  cell   during  garbage  collection
  translates  to a  membership problem  for such  grammars.  In  this
  paper, we  have shown the  undecidability of this problem and 
  circumvent the  problem by over-approximating the  grammars
  with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  show the effectiveness of liveness-based garbage collection.
\end{itemize}

As far  as we know, this  is the first work  on liveness-based garbage
collection      for      lazy       languages.       All      previous
attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
  asati14lgc, karkare06effectiveness} have  involved either imperative
or eager functional languages.
}
\subsection{Organization of the paper}

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language considered  and gives a small-step  operational semantics for
it.  The liveness  analysis for this language and  its soundness proof
is          presented          in          Section~\ref{sec:liveness}.
Section~\ref{sec:computing} describes  the formulation of  liveness as
grammars. We also give a proof  of undecidability of such grammars and
show     how     they     can     be     approximated     by     DFAs.
Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
collector,  in  particular  the  use  of  liveness  DFAs  for  garbage
collection.     We     report    our    experimental     results    in
Section~\ref{sec:experiments}    along    with   some    observations.
Section~\ref{sec:relatedwork}  discusses  previous   work  related  to
garbage  collection  and   liveness  and  Section~\ref{sec:conclusion}
discusses possible extensions and concludes the paper.
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective}, where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of closures through the $\LET$ construct.   We
  assume that  $\LET$s  in  our  language  are  non-recursive;  in  the
  expression $\LET\,\, x  \leftarrow s\,\, \IN\,\, e$,  $x$ should not
  occur in  $s$. The restriction of  \LET\ to a single  definition is
for ease  of exposition---generalization to multiple  definitions does
not add conceptual difficulties.  We further restrict each variable in
a program  to be  distinct, so that  no scope  shadowing occurs; this
simplifies reasoning.

@@

Closure  variables get  their  liveness values  through  a chain  of
dependences beginning at  a variable at an evaluation  point.  As an
example, consider the program in Figure~\ref{fig:lv-closure}.
The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
\leftarrow            (+~\py~1),\,           \py            \leftarrow
(\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed,  the chains of closures  in the
heap are a  runtime representation of these  dependences.  Since $\pz$
is evaluated  at $\psi_3$  due to  the expression  $\SRETURN~\pz$, the
demand made by the calling context(s)  of $f$ places a demand on $\pz$
which  will  impart a  liveness  to  $\px_{\pi_1}$.  Other  dependence
chains  which result  in  a liveness  for  $\px_{\pi_1}$ are  $\lbrack
\psi_1\!\!\!:\!\!\py   \leftarrow  (\length~\px_{\pi_1})\rbrack$   and
$\lbrack   \psi_2\!\!\!:\!\!\!\pw   \leftarrow  (/~\pu~\py),\,\,   \py
\leftarrow   (\length~\px_{\pi_1})\rbrack$.   The   liveness  analysis
described in this section declares the liveness of $\px_{\pi_1}$ to be
a union of  the liveness arising from these dependence  chains.  To be
safe, a garbage collection during  evaluation of $\py$ at $\psi_1$ has
to use  this liveness to  copy the heap starting from  $\px_{\pi_1}$.  However,
notice that if a garbage collection takes place while evaluating $\pz$
at $\psi_3$, it can safely consider only the liveness arising from the
dependence chain\linebreak  $\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow
(+~\py~1),\, \py \leftarrow (\length~\px_{\pi_1})\rbrack$.  The garbage
collection  scheme  described  in Section~\ref{sec:GC-scheme}  uses  a
generalization of this observation  to dynamically select a evaluation
point specific liveness in order to collect more garbage.

Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
parts. The  function $\mathit{ref}$,  takes an  application $s$  and a
demand $\sigma$ and returns the incremental liveness generated for the
free variables of $s$ due to  the application.  This will be consulted
during garbage collection  while exploring the heap  starting from the
closure  variables.  The function $\mathcal{L}$  uses  $\mathit{ref}$ to  propagate
liveness across expressions.

@@

As noted earlier,  $x.\alpha \in \Lv_y$ specifies the  liveness of the
reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$
corresponds to a closure-free path  in \heap\ starting from $\rho(x)$.
If this path is intercepted by a closure, say $(\CAR~{x_{\pi}})$, then
the liveness of further heap contents starting from $\mathit{x_{\pi}}$
is given by $\Lv_\mathit{x_{\pi}}$.

@@

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

For the inductive step, we shall show that none of the minefield steps
that involves dereferencing  results in a \bang.  These  are the steps
which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
inserted  by  the  immediately  preceding garbage  collection  or  the
garbage collection of an earlier minefield step.  However the
  demand $\sigma'$ on basis of  which an earlier collection would have
  inserted a $\bot$, would have  included the current demand $\sigma$.
  Thus, if  the earlier garbage  collection had inserted a  $\bot$, so
  would the current garbage collection.  Thus it is enough to consider
  the $\rightsquigarrow$  along with the  current GC and show  that it
  does not lead to a \bang.

@@

For the  {\sc car-bang}  rule, observe that  there are
two dereferences.  First $x$  is dereferenced to  get a
cons  cell  and then  the  head  of  the cons  cell  is
dereferenced  to  obtain  a  closure.   If  the  demand
$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
of    $x$   will    include    both   $\epsilon$    and
$\acar\epsilon$,  and  a  garbage collection with  this  liveness  will
neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
first   component  of   the   cons   cell.  Thus   both
dereferences    can take place without 
$\rightsquigarrow$ entering the   \bang\ state.
\end{proof}

@@

For  the  running  example (Figure~\ref{fig:mot-example2-a}),  $\length$  has  calls  from  $\main$  at
$\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
are $\epsilon$ and $\clazy\sigma_{\mathit{\length}}$. Thus:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the closure variables $\xl_{\mbox{$\pi$}_1}$ and $\xl_{\mbox{$\pi$}_5}$, and the
stack variables $\pl$ and $\pa$ have the following liveness in terms
of  $\sigma_{\length}$:
\begin{align*}
  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} \\
  \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
  \clazy\Df{\length}{1}{\clazy}) \sigma_{\mathit{\length}} \\
  \Lanv{\pl}{} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}} \\
  \Lanv{\pa}{} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
\end{align*}

@@

\vspace*{-2.5mm}
\section{The garbage collection scheme}
\label{sec:GC-scheme}

@@

\vspace*{-2mm}
\section{Related work} 
\label{sec:relatedwork}
The impact of  liveness on the effectiveness of  garbage collection is
investigated in~\cite{Hirzel}.   The observation is that  liveness can
significantly  impact   garbage  collection,  but  only   when  it  is
interprocedural.  As far as memory requirement is concerned, our paper
demonstrates this observation.

>>>>

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection (GC) in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis  has  two parts.   We  first  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands  on its  arguments.   This  summary is  used  to step  through
function calls  during analysis.   The concrete  demand on  a function
body  is  obtained through  a  conservative  approximation similar  to
0-CFA~\cite{Shivers:1988} that  combines the demands on  all the calls
to  the function.   The result  of the  analysis is  an annotation  of
certain program points with  deterministic finite-state automata (DFA)
capturing the  liveness of  variables at  these points.   Depending on
where GC is triggered, the collector consults a set of
automata to restrict reachability during  marking.  This results in an
increase in  the garbage reclaimed  and consequently in  fewer 
collections.

\todelete{While  this  idea has  been  shown  to  be effective  for  a
  first-order     {\em     eager}    language~\cite{asati14lgc},     a
  straightforward extension of the technique  is not possible for lazy
  languages,  where   heap-allocated  objects  may   include  closures
  (runtime   representations    of   unevaluated    expressions).    }
\added{Designing a liveness based garbage collector for lazy languages
  poses significant challenges.}  Firstly, since  data is made live by
evaluation of closures, and in lazy languages the place in the program
where  this evaluation  takes place  cannot be  statically determined,
laziness   complicates  liveness   analysis  itself.    Moreover,  for
liveness-based GC  to be effective, we  need to extend
it to closures apart from evaluated  data.  Since a closure can escape
the  scope in  which it  was  created, it  has to  carry the  liveness
information  of  its  free  variables.  As  execution  progresses  and
possible  future   uses  are   eliminated,  we  update   the  liveness
information  in a  closure with  a more  precise version.   Because of
these reasons,  the garbage collector also  becomes significantly more
complicated than a liveness-based collector for a eager language.

@@

%------------------------------------------------------------%
\subsection{Motivating example}
\label{sec:motiv}
\input{fig_motiv}   Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e. just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus if a GC takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are
\added{
\begin{itemize}
\item We propose a liveness based GC scheme for a lazy
  first order functional  language and prove its  correctness.  To the
  best  of our  knowledge, this  is the  first work  on liveness-based
  GC   for    lazy    languages.    All    previous
  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
    asati14lgc,    karkare06effectiveness}   have    involved   either
  imperative  or  eager  functional  languages.   The  novely  of  our
  approach is in  handling of  liveness of  functional values  (i.e.,
  closures), an issue that arises due to laziness.

@@

\item The  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether to  copy a  cell during GC translates  to a
  membership problem for such grammars.   In this paper, we have shown
  the undecidability  of this  problem. We  circumvent the  problem by
  over-approximating the grammars with DFAs.

\item We have implemented a garbage  collector that uses the result of
  liveness  analysis to  retain  live cells.   Our experiments  reveal
  interesting  space-time   trade-offs  in  the  engineering   of  the
  collector---for  example, updating  liveness information  carried in
  closures during  execution results in more  garbage being collected.
  Empirical results  show the effectiveness of  liveness-based GC.
\end{itemize}
}
\todelete{
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
  functional language and  prove its correctness. The 
  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
  correct with respect to the specification.   
\item The analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether   to  copy  a  cell   during  GC  translates  to a  membership problem  for such  grammars.  In  this
  paper, we  have shown the  undecidability of this problem and 
  circumvent the  problem by over-approximating the  grammars
  with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  show the effectiveness of liveness-based GC.
\end{itemize}

As far  as we know, this  is the first work  on liveness-based GC      for      lazy       languages.       All      previous
attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
  asati14lgc, karkare06effectiveness} have  involved either imperative
or eager functional languages.
}
\subsection{Organization of the paper}

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language considered  and gives a small-step  operational semantics for
it.  The liveness  analysis for this language and  its soundness proof
is          presented          in          Section~\ref{sec:liveness}.
Section~\ref{sec:computing} describes  the formulation of  liveness as
grammars. We also give a proof  of undecidability of such grammars and
show     how     they     can     be     approximated     by     DFAs.
Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
collector,  in  particular  the  use  of  liveness  DFAs  for  GC.     We     report    our    experimental     results    in
Section~\ref{sec:experiments}    along    with   some    observations.
Section~\ref{sec:relatedwork}  discusses  previous   work  related  to
GC  and   liveness  and  Section~\ref{sec:conclusion}
discusses possible extensions and concludes the paper.
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective}, where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of closures through the $\LET$ construct.   We
  assume that  $\LET$s  in  our  language  are  non-recursive;  in  the
  expression $\LET\,\, x  \leftarrow s\,\, \IN\,\, e$,  $x$ should not
  occur in  $s$. The restriction of  \LET\ to a single  definition is
for ease  of exposition---generalization to multiple  definitions does
not add conceptual difficulties.  We further restrict each variable in
a program  to be  distinct, so that  no scope  shadowing occurs; this
simplifies reasoning.

@@

Closure  variables get  their  liveness values  through  a chain  of
dependences beginning at  a variable at an evaluation  point.  As an
example, consider the program in Figure~\ref{fig:lv-closure}.
The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
\leftarrow            (+~\py~1),\,           \py            \leftarrow
(\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed,  the chains of closures  in the
heap are a  runtime representation of these  dependences.  Since $\pz$
is evaluated  at $\psi_3$  due to  the expression  $\SRETURN~\pz$, the
demand made by the calling context(s)  of $f$ places a demand on $\pz$
which  will  impart a  liveness  to  $\px_{\pi_1}$.  Other  dependence
chains  which result  in  a liveness  for  $\px_{\pi_1}$ are  $\lbrack
\psi_1\!\!\!:\!\!\py   \leftarrow  (\length~\px_{\pi_1})\rbrack$   and
$\lbrack   \psi_2\!\!\!:\!\!\!\pw   \leftarrow  (/~\pu~\py),\,\,   \py
\leftarrow   (\length~\px_{\pi_1})\rbrack$.   The   liveness  analysis
described in this section declares the liveness of $\px_{\pi_1}$ to be
a union of  the liveness arising from these dependence  chains.  To be
safe, a GC during  evaluation of $\py$ at $\psi_1$ has
to use  this liveness to  copy the heap starting from  $\px_{\pi_1}$.  However,
notice that if a GC takes place while evaluating $\pz$
at $\psi_3$, it can safely consider only the liveness arising from the
dependence chain\linebreak  $\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow
(+~\py~1),\, \py \leftarrow (\length~\px_{\pi_1})\rbrack$.  The garbage
collection  scheme  described  in Section~\ref{sec:GC-scheme}  uses  a
generalization of this observation  to dynamically select a evaluation
point specific liveness in order to collect more garbage.

Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
parts. The  function $\mathit{ref}$,  takes an  application $s$  and a
demand $\sigma$ and returns the incremental liveness generated for the
free variables of $s$ due to  the application.  This will be consulted
during GC  while exploring the heap  starting from the
closure  variables.  The function $\mathcal{L}$  uses  $\mathit{ref}$ to  propagate
liveness across expressions.

@@

As noted earlier,  $x.\alpha \in \Lv_y$ specifies the  liveness of the
reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$
corresponds to a closure-free path  in \heap\ starting from $\rho(x)$.
If this path is intercepted by a closure, say $(\CAR~{x_{\pi}})$, then
the liveness of heap contents starting from $\mathit{x_{\pi}}$
is given by $\Lv_\mathit{x_{\pi}}$.

@@

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual GC.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  GC
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

For the inductive step, we shall show that none of the minefield steps
that involves dereferencing  results in a \bang.  These  are the steps
which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
inserted  by  the  immediately  preceding garbage  collection  or  the
GC of an earlier minefield step.  However the
  demand $\sigma'$ on basis of  which an earlier collection would have
  inserted a $\bot$, would have  included the current demand $\sigma$.
  Thus, if  the earlier garbage  collection had inserted a  $\bot$, so
  would the current GC.  Thus it is enough to consider
  the $\rightsquigarrow$  along with the  current GC and show  that it
  does not lead to a \bang.

@@

For the  {\sc car-bang}  rule, observe that  there are
two dereferences.  First $x$  is dereferenced to  get a
cons  cell  and then  the  head  of  the cons  cell  is
dereferenced  to  obtain  a  closure.   If  the  demand
$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
of    $x$   will    include    both   $\epsilon$    and
$\acar\epsilon$,  and  a  GC with  this  liveness  will
neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
first   component  of   the   cons   cell.  Thus   both
dereferences    can take place without 
$\rightsquigarrow$ entering the   \bang\ state.
\end{proof}

@@

For  the  running  example (Figure~\ref{fig:mot-example2-a}),  $\length$  has  calls  from  $\main$  at
$\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
are $\epsilon$ and $\clazy\sigma_{\mathit{\length}}$. Thus:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the closure variables $\xl_{\mbox{$\pi$}_1}$ and $\xl_{\mbox{$\pi$}_5}$, and the
stack variables $\pl$ and $\pa$ have the following liveness in terms
of  $\sigma_{\length}$:
\begin{align*}
  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} 
\end{align*}
\begin{align*}
  \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
  \clazy\Df{\length}{1}{\clazy}) \sigma_{\mathit{\length}} \\
  \Lanv{\pl}{} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}} \\
  \Lanv{\pa}{} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
\end{align*}

@@

\vspace*{-2.5mm}
\section{The GC scheme}
\label{sec:GC-scheme}

@@

\vspace*{-2mm}
\section{Related work} 
\label{sec:relatedwork}
The impact of  liveness on the effectiveness of  GC is
investigated in~\cite{Hirzel}.   The observation is that  liveness can
significantly  impact   garbage  collection,  but  only   when  it  is
interprocedural.  As far as memory requirement is concerned, our paper
demonstrates this observation.
